When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-16 doc/paper/body.tex
now done with methods, except need to write out experiment table

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015
database (data available at \sloppy
\url{https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/} \sloppy),
including all Tier 1 sites with at least four years of data and
observations of the variables described in the methods section (Sect.
\ref{methods}). Sixty-six sites met these requirements, and were grouped
into nine plant functional types (PFT) according to the International
Geosphere-Biosphere Programme vegetation classification scheme
\citep{Loveland_1999}: cropland (CRO), grass (GRA), deciduous broadleaf
forest (DBF), evergreen broadleaf forest (EBF), evergreen needleleaf
forest (ENF), mixed forest (MF), closed shrub (CSH), savannah (SAV), and
woody savannah (WSA) (site locations and citations are in Fig.
\ref{map_fig} and Appendix \ref{flux_sites}, respectively).

\begin{figure}[h]
  \centering \includegraphics[trim={0 3cm 0 3cm}, clip]{./map.pdf}
  \caption{Plant functional type and location of FLUXNET2015 sites
    used in this analysis.}
  \label{map_fig}
\end{figure}

The purpose of this study is to examine ecosystem response to
atmospheric drying, focusing on the growing season. To accomplish
this, we filter and quality control the data using a similar procedure
as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data,
  according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the
  average Gross Primary Productivity (GPP) exceeds 10\% of the
  observed 95th percentile of GPP for a given site. GPP is calculated
  using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues
  with rain interception and sensor saturation at high relative
  humidity (\cite{Medlyn_2017}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime,
which is identified when downwelling shortwave radiation is greater
than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W
m$^{-2}$. To reduce the chance of sensor saturation at high relative
humidity, we remove all time steps for which VPD is less than .01 kPa,
and to reduce errors at low windspeeds we remove all periods with wind
magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed
GPP or ET are also removed, and we aggregate half hourly data to
hourly averages to reduce noise \citep{Lin_2018}. After these quality
control procedures, 400,983 upscaled hourly observations remain.

\subsection{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with
temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$),
$R_{net}$ is the net radiation minus ground heat flux, $g_a$ is
aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific
heat of air at constant pressure, $\gamma$ is the psychometric
constant, and $g_s$ is the stomatal conductance (Table
\ref{definitions}).

@@

Given eddy-covariance FLUXNET2015 data (Sect. \ref{data}), every
term in our new version of PM (Eq. (\ref{et})) is observed, except
for uWUE, which we fit by calculating its expectation, given the model
and FLUXNET2015 data.

However, eddy-covariance data are inherently noisy so we include a
measure of uncertainty in our analysis. To account for observational
error, as well as model uncertainty (e.g. temporal and spatial
variations of uWUE and $g_1$), we introduce an uncertainty parameter
$\sigma$ modifying uWUE:

% \begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R_{net} + \frac{g_a\; P}{T} \left( \frac{ c_p VPD}{R_{air}} -  \frac{\gamma c_a \sqrt{VPD} }{ R \; 1.6\; \sigma \; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})} \right) }{ \Delta + \gamma}
    \label{et_sigma}
  \end{equation}
% \end{linenomath*}

Now, from each FLUXNET2015 observation (i.e. for each hourly
observation at every time step) we can evaluate $\sigma$:

% \begin{linenomath*}
  \begin{equation}
    \sigma = - \frac{g_a \gamma c_a \sqrt{VPD} L_v P }{ \left(\text{ ET } ( \Delta + \gamma) - \Delta R_{net} - g_a \rho_a c_p VPD\right) 1.6 \; R\; T\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{VPD}})},
    \label{sigma}
  \end{equation}
% \end{linenomath*}

So, with this uncertainty analysis we can evaluate departure from
our theory in observations, as a departure of $\sigma$ from unity. The
variability of $\sigma$ across sites and time provides a measure of
uncertainty in our model, assumptions, as well as in the FLUXNET2015
observations themselves. The variability of $\sigma$ then propagates
 any uncertainty through to our partial derivative of Eq.
(\ref{et_sigma}) with respect to VPD:

% \begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; VPD} = \frac{2\; g_a \; P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -  \frac{\gamma c_a }{1.6 \; R\; \sigma \; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right) \right)
    \label{d_et}
  \end{equation}
% \end{linenomath*}

With Eq. (\ref{d_et}) we have an analytical framework for ecosystem
response to atmospheric demand perturbations with environmental
conditions held fixed. There are a few subtleties to taking the
derivative in Eq. (\ref{d_et}): $\Delta$ ($\frac{d e_{s}}{d T}$)
and $VPD$ are functionally related, so while taking the derivative we
evaluate
$\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
{\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
\Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
\frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
    fixed}}$. $RH$ and $e_s$ are assumed to be approximately
independent, which is supported by the data (not shown).

This derivation relied either implicitly or explicitly on several
assumptions. First, we assume that VPD at the leaf surface is the same
as VPD at measurement height; physically this implies that leaves are
perfectly coupled to the atmosphere. In reality, for some conditions
and plant types the leaves can become decoupled from the boundary
layer \citep{De_2017, Medlyn_2017}. Given our focus on the growing
season, which is usually characterized by relatively high insolation
inducing instability and convective boundary layers, we would expect
the surface to be generally well coupled. An additional assumption in
the formulation of $uWUE$ \citep{Zhou_2014, Zhou_2015} and
\citet{Medlyn_2017}'s stomatal conductance model is that direct soil
evaporation (E) contributions to ET remain small relative to
transpiration (T). This should be more true during the growing
season. The ratio of E to T may increase immediately after rainfall
events due to high soil moisture and ponding, but VPD is generally low
anyways during these times. However, some plant types allow for
systematically larger contributions of E in ET, particularly those
with sparse canopies and smaller relative amounts of transpiration. We
therefore might expect that the theory will be most applicable to
forest PFTs, which will be most strongly coupled to the boundary layer
due to larger surface roughness, and will also generally have the
highest ratios of transpiration to evaporation. Finally, for the goal
of developing PFT-wide intuition, we assume that $g_1$ and $uWUE$ are
constant in space and time. Both quantities have been shown to be well
conserved over space and time within a PFT relative to inter-PFT
variability. However, we would expect them to vary somewhat with
environmental conditions including very low soil water content,
temperature, and with intra-PFT plant-specific characteristics like
wood density for tree PFTs \citep{Lin_2015}. Given these strong
assumptions made with the goal of understanding broad, leading order
plant behaviors, we take the extra care of including and analyzing
uncertainty (Sect. \ref{testing}) and examining sensitivity to the
exponent of the VPD response (Sect. \ref{functional_form}).

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Eq. (\ref{et}) is fully defined
using commonly available weather station data and the constants
published here. This makes Eq. (\ref{et}) useful in addition to
PET in drought indices and hydrometeorological analysis for vegetated
surfaces. Equation (\ref{et}) better reflects the physics of water
exchange at the land surface and would only require fitting of uWUE
and $g_1$, and can also account for changes in CO$_2$ concentration
(assuming some relationship between $\lambda$ and [CO$_2$]), which is missing
in other drought indices such as the Palmer Drought Severity Index
(PDSI) \citep{Swann_2016, Lemordant_2016, Lemordant_2018}. Lastly, all
code and data used in this analysis, including those used to generate
the figures and tables, are publicly available at
\url{https://github.com/massma/climate\_et}.
%\hyperref[https://github.com/massma/climate\_et]{https://github.com/massma/climate_et}.

@@

\begin{equation}
  \frac{g_a \; P}{T(\Delta + \gamma)},
\end{equation}

@@

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_a }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the
first term and the second term in the sign term (Eq. (\ref{sign}))
will determine whether ET increases or decreases with increasing
VPD. If the second term is larger then plant control dominates and ET
decreases with increasing VPD. However, if the first term is larger,
then atmospheric demand dominates and ET increases with increasing
VPD.

>>>>

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948,
  Monteith_1965} estimates ET as a function of observable atmospheric
variables and surface conductances:
% \begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
% \end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with
temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$),
$R_{net}$ is the net radiation minus ground heat flux, $g_a$ is
aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific
heat of air at constant pressure, $\gamma$ is the psychometric
constant, and $g_s$ is the stomatal conductance (Table
\ref{definitions}).

@@

With Equation \ref{eq} we can take the partial derivative with respect
to VPD:

% \begin{linenomath*}
  \begin{equation}
    \frac{\partial \;  ET}{\partial \; VPD} = \frac{2\; g_a \;
      P}{T(\Delta + \gamma)}   \left(\frac{ c_p}{R_{air}} -
      \frac{\gamma c_a }{1.6 \; R\; \; \text{ uWUE }} \left(
        \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right)
    \right),
    \label{d_et}
  \end{equation}
% \end{linenomath*}

  deriving an analytical framework for ecosystem response to
  atmospheric demand perturbations with environmental conditions held
  fixed. There are a few subtleties to taking the derivative in
  Eq. (\ref{d_et}): $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are
  functionally related, so while taking the derivative we evaluate
  $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET}
  {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD}
  \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH}
  \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$
      fixed}}$. $RH$ and $e_s$ are assumed to be approximately
  independent, which is supported by data (not shown).

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefor, our
  derivation will be most applicable in times like the growing season,
  when relatively high insolation induces instability and convective
  boundary layers, and we would expect the surface to be generally
  well coupled. An additional assumption in the formulation of $uWUE$
  \citep{Zhou_2014, Zhou_2015} and \citet{Medlyn_2017}'s stomatal
  conductance model is that direct soil evaporation (E) contributions
  to ET remain small relative to transpiration (T). This should be
  more true during the growing season. The ratio of E to T may
  increase immediately after rainfall events due to high soil moisture
  and ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of E in ET, particularly those with sparse canopies
  and smaller relative amounts of transpiration. We therefore might
  expect that the theory will be most applicable to forest PFTs, which
  will be most strongly coupled to the boundary layer due to larger
  surface roughness, and will also generally have the highest ratios
  of transpiration to evaporation. Finally, we assume that $g_1$ and
  $uWUE$ are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in
  % TODO: make this a general discussion on partial derivatives and
  % what things are constant/how to interpret?
  VPD \cite{Franks_2017, Zhou_2014}. These parameters will however
  vary with plant species and characteristics \citep[e.g. wood density,
  ]{Lin_2015}, as well as environmental conditions including soil
  water content and temperature.
  % TODO: write exact parameter ranges we use, and introduce
  % scaling/sign term here?
  % \item

To account for both this variability in plant parameters and the
environmental, we will systematically analyze how the ET reponse to
VPD (Equation \ref{d_et}) varies. Equation \ref{d_et} includes a
``sign'' term  that determines the sign of the response in addition to
magnitude:

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_a }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right),
\end{equation}

and a ``scaling'' multiplying the ``sign'' term:

In the ``sign term'' most of quantities are relatively constant,
except for VPD and the plant parameters g$_1$ and uWUE. In the scaling
term, most of the terms are relatively constant with the exception of
aerodynamic conductance ($g_a$) and temperature (especially its effect
on $\Delta$). To determine the range of probable ET reponses to VPD we
will systematically vary this parameters according to Table
\ref{param_values}. Environmental parameters ($g_a$, $T$) are varied
according to ranges we would expect for growing season conditions
across a range of plant types (canopy height is a determinant $g_a$),
and plant parameters (g$_1$, uWUE) are varied according to ranges
established in previous literature \citep{Zhou_2015,
  Medlyn_2017}. Using this previous literature we can connect the
effect of varying plant parameters to specific plant types and
characteristics. Lastly, all code and data used in this analysis,
including those used to generate the figures and tables, are publicly
available at \url{https://github.com/massma/climate\_et}.
%\hyperref[https://github.com/massma/climate\_et]{https://github.com/massma/climate_et}.

@@

@@

All variables are positive, so the relative magnitude between the
first term and the second term in the sign term (Eq. (\ref{sign}))
will determine whether ET increases or decreases with increasing
VPD. If the second term is larger then plant control dominates and ET
decreases with increasing VPD. However, if the first term is larger,
then atmospheric demand dominates and ET increases with increasing
VPD.
